---
title: "STAT 3494 Data Analysis"
author: "Patrick Brogan"
date: "2022-11-28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(lme4)
library(plyr)
library(dplyr)
library(car)
library(MASS)
library(lmtest)
```

## Rounding

```{r}
round_df <- function(x, digits) {
    # round all numeric variables
    # x: data frame 
    # digits: number of digits to round
    numeric_columns <- sapply(x, mode) == 'numeric'
    x[numeric_columns] <-  round(x[numeric_columns], digits)
    x
}
```

## Data Entry

### Asthma

```{r}
asthma.sex <- data.frame(Asthma_Sex)
colnames(asthma.sex)
colnames(asthma.sex)[3] = "SpatialDimValueCode"
colnames(asthma.sex)[5] = "Period"
asthma <- merge(PM.data, asthma.sex, by=c("Period","SpatialDimValueCode"))
GDP <- data.frame(GDP_Per_Capita)
colnames(GDP)[2] = "SpatialDimValueCode"
total.asthma <- merge(asthma, GDP, by=c("Period","SpatialDimValueCode"))
asthma.data <- transform(total.asthma, Value_cs=scale(Value), GDP_cs=scale(GDP.per.capita), Period_cs=scale(Period))
```

### Rheumatic Heart Disease

```{r}
rheum.sex <- data.frame(Rheum_Sex)
colnames(rheum.sex)
colnames(rheum.sex)[3] = "SpatialDimValueCode"
colnames(rheum.sex)[5] = "Period"
rheum <- merge(PM.data, rheum.sex, by=c("Period","SpatialDimValueCode"))
total.rheum <- merge(rheum, GDP, by=c("Period","SpatialDimValueCode"))
rheum.data <- transform(total.rheum, Value_cs=scale(Value), GDP_cs=scale(GDP.per.capita), Period_cs=scale(Period))
```

### Hypertensive Heart Disease

```{r}
hyper.sex <- data.frame(Hyper_Sex)
colnames(hyper.sex)
colnames(hyper.sex)[3] = "SpatialDimValueCode"
colnames(hyper.sex)[5] = "Period"
hyper <- merge(PM.data, hyper.sex, by=c("Period","SpatialDimValueCode"))
total.hyper <- merge(hyper, GDP, by=c("Period","SpatialDimValueCode"))
hyper.data <- transform(total.hyper, Value_cs=scale(Value), GDP_cs=scale(GDP.per.capita), Period_cs=scale(Period))
```

## Control for sex, time trends and GDP per capita

### Asthma

```{r}
total.asthma.pos <- subset(asthma.data, Age.standardized.death.rate.per.100.000.standard.population>0)
model.asthma <- lmer(Age.standardized.death.rate.per.100.000.standard.population
                     ~ Value_cs + GDP_cs + Period_cs + Sex + (0 + GDP_cs|Value_cs)
                        + (0 + Sex|Value_cs) + (0 + Period_cs|Value_cs), total.asthma.pos)
asthma0 <- lm(Age.standardized.death.rate.per.100.000.standard.population
                    ~ Value*GDP.per.capita*Period*Sex, total.asthma)
```

### Rheumatic Heart Disease

```{r}
total.rheum.pos <- subset(rheum.data, Age.standardized.death.rate.per.100.000.standard.population>0)
model.rheum <- lmer(Age.standardized.death.rate.per.100.000.standard.population
                     ~ Value_cs + GDP_cs + Period_cs + Sex + (0 + GDP_cs|Value_cs)
                        + (0 + Sex|Value_cs) + (0 + Period_cs|Value_cs), total.rheum.pos)
rheum0 <- lm(Age.standardized.death.rate.per.100.000.standard.population
                    ~ Value*GDP.per.capita*Period*Sex, total.rheum)
```

### Hypertensive Heart Disease

```{r}
total.hyper.pos <- subset(hyper.data, Age.standardized.death.rate.per.100.000.standard.population>0)
model.hyper <- lmer(Age.standardized.death.rate.per.100.000.standard.population
                     ~ Value_cs + GDP_cs + Period_cs + Sex + (0 + GDP_cs|Value_cs)
                        + (0 + Sex|Value_cs) + (0 + Period_cs|Value_cs), total.hyper.pos)
hyper0 <- lm(Age.standardized.death.rate.per.100.000.standard.population
                    ~ Value*GDP.per.capita*Period*Sex, total.hyper.pos)
```

## Checking Normality

### Asthma

```{r}
summ.asthma <- summary(model.asthma)
shapiro.test(summ.asthma$residuals)
asthma.data <- na.omit(total.asthma)
qqnorm(summ.asthma$residuals)
qqline(summ.asthma$residuals)
boxcox(hyper0, lambda = seq(-1, 1, length = 10))
powerTransform(hyper0, family="bcPower")
asthma.power <- 0.02754374
model.asthma.trans <- lmer(Age.standardized.death.rate.per.100.000.standard.population^asthma.power
                    ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value)
                    + (0 + Sex|Value) + (0 + Period|Value), total.asthma.pos)
summary(model.asthma.trans)
anova(model.asthma.trans)
coef.asthma <- data.frame(coef(summary(model.asthma.trans)))
coef.asthma$p.value <- 2 * (1 - pnorm(abs(coef.asthma$t.value)))
coef.asthma
```

We can reject the null hypothesis and conclude that the normality assumption is violated.Therefore, a power transform is required.

### Rheumatic Heart Disease

```{r}
summ.rheum <- summary(model.rheum)
shapiro.test(summ.rheum$residuals)
rheum.data <- na.omit(total.rheum)
qqnorm(summ.rheum$residuals)
qqline(summ.rheum$residuals)
rheum.power <- 0.02754374
model.rheum.trans <- lmer(Age.standardized.death.rate.per.100.000.standard.population^hyper.power
                    ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value)
                    + (0 + Sex|Value) + (0 + Period|Value), total.rheum.pos)
summary(model.hyper.trans)
anova(model.hyper.trans)
coef.rheum <- data.frame(coef(summary(model.hyper.trans)))
coef.rheum$p.value <- 2 * (1 - pnorm(abs(coef.rheum$t.value)))
coef.rheum
```

We can reject the null hypothesis and conclude that the normality assumption is violated. Therefore, a power transform is required.

### Hypertensive Heart Disease

```{r}
summ.hyper <- summary(model.hyper)
shapiro.test(summ.hyper$residuals)
hyper.data <- na.omit(total.hyper)
qqnorm(summ.hyper$residuals)
qqline(summ.hyper$residuals)
boxcox(hyper0, lambda = seq(-1, 1, length = 10))
powerTransform(hyper0, family="bcPower")
hyper.power <- 0.02754374
model.hyper.trans <- lmer(Age.standardized.death.rate.per.100.000.standard.population^hyper.power
                    ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value)
                    + (0 + Sex|Value) + (0 + Period|Value), total.hyper.pos)
summary(model.hyper.trans)
anova(model.hyper.trans)
coef.hyper <- data.frame(coef(summary(model.hyper.trans)))
coef.hyper$p.value <- 2 * (1 - pnorm(abs(coef.hyper$t.value)))
round_df(coef.hyper, 4)
```

We can reject the null hypothesis and conclude that the normality assumption is violated. Therefore, a power transform is required.

## Checking Equal Variance

### Asthma

```{r}
bartlett.test((total.asthma$Age.standardized.death.rate.per.100.000.standard.population)^asthma.power ~ total.asthma$Value)
bartlett.test((total.asthma$Age.standardized.death.rate.per.100.000.standard.population)^asthma.power ~ total.asthma$GDP.per.capita)
bartlett.test((total.asthma$Age.standardized.death.rate.per.100.000.standard.population)^asthma.power ~ total.asthma$Period)
summ.asthma.t <- summary(model.asthma.trans)
fit.asthma <- fitted(model.asthma.trans)
a.residuals <- summ.asthma.t$residuals
asthma.resfit <- data.frame(a.residuals, fit.asthma)
asthma.resfit.lm <- lm(abs(a.residuals) ~ fit.asthma, asthma.resfit)
wt.asthma <- 1/asthma.resfit.lm$fitted.values^2
length(total.asthma.pos$Period)
total.asthma.pos <- subset(total.asthma.pos, GDP.per.capita != "NA")
total.asthma.pos <- cbind(total.asthma.pos, wt.asthma)
wls.asthma <- lmer(Age.standardized.death.rate.per.100.000.standard.population^asthma.power
                    ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value)
                    + (0 + Sex|Value) + (0 + Period|Value), total.asthma.pos, weights = wt.asthma)
summary(wls.asthma)
```

### RHD

```{r}
bartlett.test((total.rheum$Age.standardized.death.rate.per.100.000.standard.population)^rheum.power ~ total.rheum$Value)
bartlett.test((total.rheum$Age.standardized.death.rate.per.100.000.standard.population)^rheum.power ~ total.rheum$GDP.per.capita)
bartlett.test((total.rheum$Age.standardized.death.rate.per.100.000.standard.population)^rheum.power ~ total.rheum$Period)
summ.rheum.t <- summary(model.rheum.trans)
fit.rheum <- fitted(model.rheum.trans)
rh.residuals <- summ.rheum.t$residuals
rheum.resfit <- data.frame(rh.residuals, fit.rheum)
rheum.resfit.lm <- lm(abs(rh.residuals) ~ fit.rheum, rheum.resfit)
wt.rheum <- 1/(rheum.resfit.lm$fitted.values)^2
total.rheum.pos <- subset(total.rheum.pos, GDP.per.capita != "NA")
total.rheum.pos <- cbind(total.rheum.pos, wt.rheum)
wls.rheum <- lmer(Age.standardized.death.rate.per.100.000.standard.population^rheum.power
                    ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value)
                    + (0 + Sex|Value) + (0 + Period|Value), total.rheum.pos, weights = wt.rheum)
summary(wls.rheum)
length(wt.rheum)
```

### HHD

```{r}
bartlett.test((total.hyper$Age.standardized.death.rate.per.100.000.standard.population)^hyper.power ~ total.hyper$Value, data = total.hyper)
bartlett.test((total.hyper$Age.standardized.death.rate.per.100.000.standard.population)^hyper.power ~ total.hyper$GDP.per.capita)
bartlett.test((total.hyper$Age.standardized.death.rate.per.100.000.standard.population)^hyper.power ~ total.hyper$Period)
summ.hyper.t <- summary(model.hyper.trans)
fit.hyper <- fitted(model.hyper.trans)
hy.residuals <- summ.hyper.t$residuals
hyper.resfit <- data.frame(hy.residuals, fit.hyper)
hyper.resfit.lm <- lm(abs(hy.residuals) ~ fit.hyper, hyper.resfit)
wt.hyper <- 1/(hyper.resfit.lm$fitted.values)^2
total.hyper.pos <- subset(total.hyper.pos, GDP.per.capita != "NA")
total.hyper.pos <- subset(total.hyper.pos, Age.standardized.death.rate.per.100.000.standard.population != "NA")
total.hyper.pos <- subset(total.hyper.pos, Sex != "NA")
total.hyper.pos <- subset(total.hyper.pos, Period != "NA")
total.hyper.pos <- subset(total.hyper.pos, Value != "NA")
total.hyper.pos <- cbind(total.hyper.pos, wt.hyper)

wls.hyper <- lmer(Age.standardized.death.rate.per.100.000.standard.population^rheum.power
                    ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value)
                    + (0 + Sex|Value) + (0 + Period|Value), total.hyper.pos, weights = wt.hyper)
summary(wls.hyper)
```

## Linearity

### Asthma

```{r}
total.asthma.pos$Death.rate.transform <- (total.asthma.pos$Age.standardized.death.rate.per.100.000.standard.population)^asthma.power
scatterplotmatrix <- scatterplotMatrix(~ Death.rate.transform + Value + GDP.per.capita + Period , total.asthma.pos, spread=FALSE, smoother.args=list(lty=2),
main="Scatter Plot Matrix")
scatterplotmatrix
```

### RHD

```{r}
total.rheum.pos$Death.rate.transform <- (total.rheum.pos$Age.standardized.death.rate.per.100.000.standard.population)^rheum.power
length(total.rheum.pos$Death.rate.transform)
scatterplotmatrix <- scatterplotMatrix(~ Death.rate.transform + Value + GDP.per.capita + Period , total.rheum.pos, spread=FALSE, smoother.args=list(lty=2),
main="Scatter Plot Matrix")
scatterplotmatrix
```

### HHD

```{r}
total.hyper.pos$Death.rate.transform <- (total.hyper.pos$Age.standardized.death.rate.per.100.000.standard.population)^rheum.power
scatterplotmatrix <- scatterplotMatrix(~ Death.rate.transform + Value + GDP.per.capita + Period , total.hyper.pos, spread=FALSE, smoother.args=list(lty=2),
main="Scatter Plot Matrix")
scatterplotmatrix
```


## Influential Observations

```{r}
influencePlot(wls.asthma, id=list(method="identify"), main="Influence Plot",
sub="Circle size is proportial to Cook's Distance" )
influencePlot(wls.rheum, id=list(method="identify"), main="Influence Plot",
sub="Circle size is proportial to Cook's Distance" )
influencePlot(wls.hyper, id=list(method="identify"), main="Influence Plot",
sub="Circle size is proportial to Cook's Distance" )
outlierTest(wls.asthma)
outlierTest(wls.rheum)
outlierTest(wls.hyper)
new.asthma <- total.asthma.pos[-959, ]
new.rheum <- total.rheum.pos[-c(750, 1244, 716, 715, 636), ]
new.hyper <- total.hyper.pos[-c(372, 1244), ]

```
## Results

### Asthma

```{r}
asthma.final <- lmer(Age.standardized.death.rate.per.100.000.standard.population^asthma.power ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value) + (0 + Sex|Value) + (0 + Period|Value), new.asthma, weights = wt.asthma)

anova(asthma.final)
summ.asthma.final <- summary(asthma.final)
coeff.asthma <- data.frame(summ.asthma.final$coefficients)
coeff.asthma$p.value <- 2*(1-pnorm(abs(coeff.asthma$t.value)))
coeff.asthma

```

### RHD 

```{r}
wt.rheum1 <- wt.rheum[-c(750, 1244, 716, 715, 636)]

rheum.final <- lmer(Age.standardized.death.rate.per.100.000.standard.population^asthma.power ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value) + (0 + Sex|Value) + (0 + Period|Value), new.rheum, weights = wt.rheum1)

anova(rheum.final)
summ.rheum.final <- summary(rheum.final)
coeff.rheum <- data.frame(summ.rheum.final$coefficients)
coeff.rheum$p.value <- 2*(1-pnorm(abs(coeff.rheum$t.value)))
pnorm(coeff.rheum$t.value)
coeff.rheum
```

### HHD

```{r}
wt.hyper1 <- wt.hyper[-c(372, 1244)]

hyper.final <- lmer(Age.standardized.death.rate.per.100.000.standard.population^asthma.power ~ Value + GDP.per.capita + Period + Sex + (0 + GDP.per.capita|Value) + (0 + Sex|Value) + (0 + Period|Value), new.hyper, weights = wt.hyper1)

anova(hyper.final)
summ.hyper.final <- summary(hyper.final)
coeff.hyper <- data.frame(summ.hyper.final$coefficients)
coeff.hyper$p.value <- 2*(1-pnorm(abs(coeff.hyper$t.value)))
coeff.hyper
```



